A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
more

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We introduce a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies observationally equal, possibly cyclic values.
The procedure is complete for the universal fragment and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
modern SMT solver. An evaluation based on problems generated from Isabelle
theories demonstrates the potential of the procedure for both proving and model
finding.
\end{abstract}

@@

Freely generated datatypes are widely used both in functional programs and in
logical specifications. They can be implemented efficiently and enjoy nice
properties that can be exploited in automated reasoners. Although they have
their uses in mathematics, they are especially prominent in computer science
applications.

However, because datatype values correspond to ground terms, they can only
represent finite objects. For example, the datatype of natural numbers
constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.
To represent infinite values such as $\const{S}(\const{S}(\ldots))$, the natural
choice is to turn to codatatypes, the coinductive dual of datatypes.

  * are sometimes perceived as esoteric
  * but they have their uses where they are nearly irreplacable
  * for example, CompCert and JinjaThreads

  * codatatype approach unifies the finite and infinite case; it is often more
    convenient to have
      codatatype enat = EZ | ES enat
    than
      datatype enat = EZ | ES enat | EI
    with $\infty$ represented by $ES(ES(\ldots))$. Notice that adding
    $\infty$, as expected

  * also can be understood very easily:
    intuitively, they consist of all ground constructor terms, finite and infinite
    and only those
  * one implication is that if $m$ equals $ES(n)$ and $n$ equals $ES(m)$, necessarily
    $m$ and $n$ must be equal.

>>>>

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We introduce a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies observationally equal, possibly cyclic values.
The procedure is complete for ground problems and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
modern SMT solver. An evaluation based on problems generated from Isabelle
theories demonstrates the potential of the procedure for both proving and model
finding.
\end{abstract}

@@

Freely generated datatypes are widely used both in functional programs and in
logical specifications. They can be implemented efficiently and enjoy nice
properties that can be exploited in automated reasoners. Although they have
their uses in mathematics, they are especially prominent in computer science
applications.
%
However, because datatype values correspond to ground terms, they can only
represent finite objects. For example, the datatype of natural numbers
constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.

To represent infinite values such as $\const{S}(\const{S}(\ldots))$, a natural
choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have a role to
play in computer science. The verified C compiler CompCert \cite{leroy-2009},
the verified Java compiler JinjaThreads \cite{lochbihler-2010-jinja}, and the
formalized Java memory model \cite{lochbihler-2014-jmm} all rely on
codatatypes to capture infinite processes.

Codatatypes, like datatypes, are freely generated by their constructors. But
whereas datatype values are finite, for codatatypes infinite constructor terms
also denote values. Intuitively, a codatatype consists of all finite and
infinite ground constructor terms, and only those. For example, the
specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{EZ} \,\mid\, \const{ES}(\ty{enat})\]
%
(expressed in a notation inspired by typed functional programming languages)
introduces a type that aptly models the natural numbers extended with an
infinite value $\infty = \const{ES}(\const{ES}(\const{ES}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{EZ} \,\mid\, \const{ES}(\ty{enat}) \,\mid\, \const{EInf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Equations such as $\const{ES}(\infty) \teq \infty$ hold without any additional
work, because both sides of the equality share the same infinite expansion
$\const{ES}(\const{ES}(\const{ES}(\ldots)))$.

%  * one implication is that if $m$ equals $ES(n)$ and $n$ equals $ES(m)$, necessarily
%    $m$ and $n$ must be equal.
